---
title: "Practicum"
author: "Pearl Naomi Munnangi"
date: "2023-06-27"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(dplyr)
library(ggplot2)
```

# 1. Data Loading

```{r,echo = TRUE}

# Load the downloaded datasets into separate dataframes and observe the dimensions
gt1 <- read_excel("G&T Results 2017-18.xlsx")
gt2 <- read_excel("G&T Results 2018-19.xlsx")

dim(gt1)
dim(gt2)

```

The two files are loaded into r dataframes, there are 117 rows and 12 columsns in 2017-18 dataset and there are 100 rows and 14 columns in 2018-19 dataset



```{r}
colnames(gt1)
colnames(gt2)

```

# 2. Data Combining
```{r}
# Remove the the Extra columns from gt2 
gt2 <- gt2[ , -c(13:14)]

# Combine the two data frames
gt <- rbind(gt1, gt2)
dim(gt)
```
Now the dataset has 217 rows and 12 columns
```{r}
str(gt)
```


# 3. Data Cleaning
#a. Missing values
```{r}
# Find the missing value counts in each feature
sapply(gt, function(x) sum(is.na(x)))

```
School preferences, School assigned and will you enroll there features are having missing values more than 10 % hence those columns/features ar excluded

```{r}
# Drop the columns with high missisng values
gt <- select(gt,-c("School Preferences","School Assigned","Will you enroll there?"))
dim(gt)
```


```{r}
#Now the rows with the mising values will be ommitted
gt <- na.omit(gt)
dim(gt)

```
Now the dataset has 193 non missing rows with 9 features 


#b. Errors

```{r}
# Some columns has score in x/y format. Convert that into numeric
select_first <- function(x) {
  sapply(strsplit(as.character(x), "/"), `[`, 1)
}

gt$"OLSAT Verbal Score" <- as.numeric(select_first(gt$"OLSAT Verbal Score"))
gt$"NNAT Non Verbal Raw Score" <- as.numeric(select_first(gt$"NNAT Non Verbal Raw Score"))

```

```{r}
# Convert other numeric variables to num from chr
gt$"OLSAT Verbal Percentile" <- as.numeric(gt$"OLSAT Verbal Percentile")
gt$"NNAT Non Verbal Percentile" <- as.numeric(gt$"NNAT Non Verbal Percentile")

str(gt)
```

```{r}
# Drop the NAs due coecion
gt <- na.omit(gt)
dim(gt)
```
Finally the dataset has 182 rows and 9 columns

```{r}
# Find the unique values in Birth months column
unique_months <- unique(gt$`Birth Month`)
print(unique_months)

```
```{r}
# there are some numeric months so replace them with month name
month.names <- c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December")

# Replace the numeric values with month names
for(i in 1:12){
  gt$`Birth Month` <- replace(gt$`Birth Month`, gt$`Birth Month` == as.character(i), month.names[i])
}

# Check the result
unique(gt$`Birth Month`)

```

```{r}
# There are some instances with wrong spelleings for febrauary and september. 
#Correct the spelling of "Febrauary" and "september"
gt$`Birth Month` <- sub("Febrauary", "February", gt$`Birth Month`)
gt$`Birth Month` <- sub("september", "September", gt$`Birth Month`)
```


# 4.Data Plottng
#a Box plot
```{r}
# Boxplot
ggplot(gt, aes(x = "NNAT Non Verbal Raw Score", y = `NNAT Non Verbal Raw Score`)) +
  geom_boxplot(fill = "#0073C2", color = "#e9ecef", alpha = 0.9) +
  labs(title="Boxplot of NNAT Non Verbal Raw Score",
       x="",
       y="NNAT Non Verbal Raw Score")

```

#b Scatter plot
```{r}
# Scatter plot
ggplot(gt, aes(x = `OLSAT Verbal Score`, y = `NNAT Non Verbal Raw Score`)) +
  geom_point(alpha = 0.6, color = "red") +
    labs(title="Scatter plot of OLSAT Verbal Score vs. NNAT Non Verbal Raw Score",
       x="OLSAT Verbal Score",
       y="NNAT Non Verbal Raw Score")
```
#c  Bar plot
```{r}
ggplot(gt, aes(x = `Birth Month`)) +
  geom_bar(fill = "lightgreen", color = "lightblue", alpha = 0.9) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="Bar Chart of Birth Month Counts",
       x="Birth Month",
       y="Count")
```

